---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)

```

## Package `dplyr`

A package with functions to manipulate *dataframes*. 

\


Some basic `dplyr` functions:

- `select()`  $\qquad$   to subset variables (columns) of a dataframe 

- `filter()`  $\qquad$   to subset observations (rows) of a dataframe  

- `mutate()`  $\qquad$   to compute new variables (or redefine existing ones)

- `rename()`  $\qquad$   to change the name of variables  

- `arrange()` $\qquad$   to sort by one or more vars


\

Load `dplyr` to make its functions available: 

```{r}
library(dplyr)
```


\

Package `dplyr` is part of the [tidyverse](https://www.tidyverse.org/), a very useful collection of packages, and a package itself. 

You can also load `dplyr` (and other packages of the tidyverse) by loading the `tidyverse` package:

```{r}
library(tidyverse)
```

