---
title: "Subsetting variables with `select()`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)

```

## General syntax

`select(dataframe, var_selection)`

\

- First argument:  dataframe to select vars from  
- Further arguments:  expression(s) indicating vars to be selected (comma-separated if more than one)  

- Result:  dataframe with selected vars only (and all rows)


## List variables (comma separated)

```{r}
library(dplyr)
d <- rio::import("data/hta.xlsx", sheet = "data") 

select(d, pid, region, age, sex)  %>%  head()

```

\

`%>% head()`  is to print only the first 6 rows (more on this later).



## Range of (consecutive) variables

```{r }
select(d, pid:sex)  %>%  head() 
```


## Exclude a variable

```{r }
select(d, pid:sex, -data_xtract_dt) %>%  head() 
```


## Helper functions  

To write expressions indicating the vars to be selected

Based on a part of the variable name  
- `contains()`  
- `starts_with()`  
- `ends_with()`  
- `num_range()`   
    
Based on a character vector of variable names  
- `all_of()`  
- `any_of()`  
    
- ... 

## Parts of var name

```{r }
select(d, pid, contains("dx"))  %>% head(2)                    # vars containing "dx"
```

```{r }
select(d, pid, starts_with("dx"))  %>% head(2)                 # vars starting with "dx"
```

```{r}
select(d, ends_with("dt"))   %>% head(2)                       # vars ending with "dt" 
```

## Parts of var name 

`num_range(fixed_part, range_for_suffix)`

\

```{r}
select(d, num_range("sbp_v", 1:3)) %>% head(2)                 # SBP repetitions 1 to 3
select(d, num_range("dbp_v", 2:5)) %>% head(2)                 # DBP repetitions 2 to 5
```


## Variable names in a character vector

```{r}
demovars <- c("pid", "age", "sex", "region")                   # character vector
demovars

select(d, any_of(demovars)) %>% head()

```


## Exercises

From dataframe `d`:

1. Select `pid` and all BP vars using `contains()`, and save as `bp` 
1. Select all variables excluding all BP variables, and save as `non_bp`
1. Select `pid` and all diagnostics using `starts_with()`, and print first 6 rows
1. Select `pid` and all dates, and print first 6 rows
1. Select `pid` and all treatment variables (`lmr` to `other`), and print first 6 rows
