---
title: "Subsetting observations"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)

library(dplyr)
d <- rio::import("data/hta.xlsx", sheet = "data") 

```


## Several `dplyr` functions

--------------    ----------------------------------------------------------------
`slice()`         based on position of rows
`slice_min()`     based lowest values in a variable
`slice_max()`     based highest values in a variable
`sample_n()`      random selection
`filter()`        based on a logical expression (*)
--------------    ----------------------------------------------------------------

\

(*) something that resolves to a logical value (TRUE or FALSE)

- relational operators:  $\qquad$ < &nbsp; , &nbsp; > &nbsp; , &nbsp; <= &nbsp; , &nbsp; >= &nbsp; , &nbsp; == &nbsp; , &nbsp; !=

- may be combined with logical operators:  $\qquad$  & &nbsp; , &nbsp; | &nbsp; , &nbsp; !


## Based on a logical expression

```{r}
demo <- select(d, pid, sex, age, region)             # easier to see with fewer vars

```

\

`filter(dataframe, logical_condition)` 

```{r}
filter(demo, age > 90)                               # simple condition
```


## Based on a logical expression

`filter(dataframe, logical_condition)` 

```{r}
filter(demo, sex == 1 & age > 85)                    # complex condition
```



## Based on a logical expression

`filter(dataframe, logical_condition)` 

```{r}
filter(demo, is.na(sex))                             # sex is missing
```


## Based on a logical expression

`filter(dataframe, logical_condition)` 

```{r}
filter(demo, !is.na(sex))                            # sex is NOT missing
```

## Based on position of rows

`slice(dataframe, row_selection)` 

```{r}
slice(demo, 1:10)
```

`row_selection` can be any numeric vector of integers

## Lowest or highest values in a variable

`slice_min(dataframe, variable, n)`   
`slice_max(dataframe, variable, n)`

```{r}
slice_min(demo, age, n = 3)                   # 3 obs with lowest age
slice_max(demo, age, n = 3)                   # 3 obs with highest age
```

## Random sample

`sample_n(dataframe, n)`   

```{r}
set.seed(1)                                   # for reproducibility
sample_n(demo, 10)                            # random selection of 10 obs
```

## Exercises

1. Subset current smokers
1. Subset current smokers or ex-smokers
1. Same, but younger than 65
1. Subset the 10 youngest patients
1. Get a random sample of size 100