---
title: "Chaining operations"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)

library(dplyr)
d <- rio::import("data/hta.xlsx", sheet = "data") 

```


## Chaining operations with `%>%`

Suppose we want to select demographic data for women only.

```{r results='hide'}
demo <- select(d, pid, age, sex)                       # option 1
filter(demo, sex == 2)

filter(select(d, pid, age, sex), sex == 2)             # option 2

d %>%                                                  # option 3
  select(pid, age, sex) %>% 
  filter(sex == 2)

```

Read $\quad$ `%>%` $\quad$ as $\quad$ *then*:

take `d` $\quad$ *then*  
$\quad$ `select` pid, age, sex $\quad$ *then*  
$\quad$ `filter` females
  
## The *pipe* operator

`%>%` is called the *pipe* operator

The *first* argument of a function can be extracted and put *before* the pipe:

<center>
`x %>% f()`  $\qquad \equiv \qquad$  `f(x)`
</center>

\

If a function takes several arguments, *the first one* can be put before the pipe:

<center>
`x %>% f(y, z, ...)`  $\qquad \equiv \qquad$  `f(x, y, z, ...)`
</center>

## Examples

```{r}
round(pi, 4)
pi %>% round(4)
log(10) 
10 %>% log()
```

## Examples

Most `dplyr` functions take a dataframe as first argument, and produce a dataframe as result. Therefore, they can be chained.

```{r}
d %>%                                   # take demo, THEN           
  select(pid, age, sex) %>%             # select pid, age and sex, THEN
  filter(sex == 2) %>%                  # filter females, THEN
  head()                                # print 6 first rows only
```

You can chain *any* number of operations

## Exercises

Use the pipe to chain operations needed in each case:

1. Subset patients from Catalunya and select variables `pid`, `age`, and `sex`
1. Could you do the same inverting the order of operations? 
    + first select variables, 
    + then filter cases from Catalunya
  
1. Same, but sort cases by age; hint: see `?arrange`
1. Same, but sorting by decreasing age; hint: see `?arrange`, second example.
1. Same but sorting by sex and decreasing age
