---
title: "Computing variables with `mutate()`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)

```


## General syntax

`mutate(dataframe, varname = expression, varname = expression, ...)`

Common cases:

- Define factors for categorical vars
- Compute numeric var with a formula  
- Conditional assignment  
- Categorize numeric var
- Recoding 

\

```{r}
library(dplyr)
d <- rio::import("data/hta.xlsx", sheet = "data")
```



## Define factors

```{r}
d %>% 
  mutate(sex = factor(sex, levels = 1:2, labels = c("male", "female")),
         smoke = factor(smoke, 
                        levels = 1:4, 
                        labels = c("current smoker", "never smoker", 
                                   "ex-smoker", "unknown")
                        )
         ) %>% 
  select(sex, smoke) %>% head()

```


## Compute numeric var with a formula

```{r}
d %>% 
  mutate(bmi = weight / height^2) %>%                       # body mass index
  select(pid, weight, height, bmi) %>% head(3)               

d %>% 
  mutate(bmi = round(weight / height^2, 1)) %>%              # rounding to 1st decimal
    select(pid, weight, height, bmi) %>% head(3)       
```


## Conditional assignment (simple)

`if_else(logical_condition, value_if_TRUE, value_if_FALSE)`

\

Obesity is defined as a body mass index of $\hspace{4 mm}30 \hspace{2 mm} kg/m^2 \hspace{4 mm}$ or higher

```{r}
d %>% 
  mutate(bmi = round(weight / height^2, 1),
         obesity = if_else(bmi >= 30, "yes", "no")) %>%
  select(pid, weight, height, bmi, obesity) %>% head()     
```


## Conditional assignment (complex)

`case_when(logical_condition ~ value,`  
&nbsp; &nbsp;&nbsp;&nbsp;&nbsp;&nbsp; &nbsp; &nbsp;&nbsp;&nbsp;&nbsp;&nbsp; &nbsp;&nbsp;&nbsp;&nbsp; `logical_condition ~ value, ...)`


```{r}
d %>% 
  mutate(x = case_when(dx_dm == 1 & sex == 1 ~ "diabetic male",
                       dx_dm == 1 & sex == 2 ~ "diabetic female",
                       dx_dm == 2 & sex == 1 ~ "non-diabetic male",
                       dx_dm == 2 & sex == 2 ~ "non-diabetic female")) %>% 
           select(dx_dm, sex, x) %>% head()
```



## Categorize a numeric var

`cut(numeric_vector, breaks, ...)`

\

```{r}

d %>% 
  mutate(bmi = round(weight / height^2, 1),
         who_ns = cut(bmi, 
                      breaks = c(0, 18.5, 25, 30, 35, 40, Inf))) %>% 
  select(weight, height, bmi, who_ns) %>% head(3)
```

`(25,30]`  means  *left-open* and *right-closed* interval: 30 is included, 25 is not.  

What if we want the opposite?


## Categorize a numeric var: *right-open* intervals

```{r}

d %>% 
  mutate(bmi = round(weight / height^2, 1),
         who_ns = cut(bmi, 
                      breaks = c(0, 18.5, 25, 30, 35, 40, Inf),
                      right = FALSE)) %>% 
  select(weight, height, bmi, who_ns) %>% head()
```

What if we want custom labels?


## Categorize a numeric var: custom labels

```{r}

d <- d %>% 
  mutate(bmi = round(weight / height^2, 1),
         who_ns = cut(bmi, 
                      breaks = c(0, 18.5, 25, 30, 35, 40, Inf),
                      right = FALSE,
                      labels = c("Underweight", "Normal weight",
                                 "Overweight", "Obesity class I",
                                 "Obesity class II", "Obesity class III")
                      ))  
d %>% select(weight, height, bmi, who_ns) %>% head()
```



## Exercises

Create dataframe `demo` with variables `pid`, `region`, `sex`, `age`, `smoke`, `height` and `weight`, and:

1. compute the body mass index, rounded to first decimal.
1. create a logical indicator for obesity; hint: use `if_else(..., TRUE, FALSE)`.
1. create variable `age_group` to group ages in the following categories, labeling them as "young", "mature", "old", respectively: 
    + less than 40 
    + 40 to less than 65 
    + 65 or more  

1. redefine `smoke` as "current or ex-smoker", "never smoker", or "unknown".
1. define factors for all categorical variables.




 


