---
title: "Strings"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)


```


## Common problems

```{r}
library(tidyverse)

antec <- c("Arterial hypertension", "arterial hypertension", "ARTERIAL HYPERTENSION",
       "arterial   hypertension ", "arterial hypertension ",  "Pneumonia", 
       "pneumonia", " pneumonia")

data.frame(antec) %>% count(antec)
```


## Whitespaces (ws)

```{r}
str_trim(antec)                            # removes leadin/trailing ws
```

\

```{r}
str_squish(antec)                          # also removes excessive ws between words
```


## Capitalization

```{r}
str_to_upper(antec)                        # converts to uppercase
```

\

```{r}
str_to_sentence(antec)                     # capitalizes first letter
```


## Several operations

```{r}

str_to_sentence(str_squish(antec))
antec %>% str_squish() %>% str_to_sentence()
antec %>% str_squish() %>% str_to_sentence() %>% unique()
```


## Substrings

```{r}
x <- c("FEGA0720525000", "ROMA1680913000", "PEJI0590503")
```

\

```{r}
str_sub(x, start = 5, end = 5)          # extracts sex code
```

\

```{r}
str_sub(x, start = 6, end = 11)         # extracts birthdate
```


## Matching patterns

```{r}
x <- c("diabetes mellitus",  "asthma", "bronchial asthma + DM.")
```

\

```{r}
str_detect(x, pattern = "asthma")                                      # detect pattern
```

\

```{r}
str_replace(x, pattern = "DM", replacement = "diabetes mellitus")      # detect and replace
```



## Matching patterns: care with special characters!

```{r}
x <- c("diabetes mellitus",  "asthma", "bronchial asthma + DM.")
```

\

```{r}
str_detect(x, ".")                                # unexpected result !
```

\

```{r}
str_detect(x, fixed("."))                         # OK
```


## Other useful functions

```{r}
x <- c("asthma", "diabetes mellitus", "preunomia")
length(x)
```


```{r}
str_length(x)                                     # counts number of characters 
```

\

```{r}
str_flatten(x, collapse = ", ")
```

## Other useful functions

```{r}
age <-  c(35, 48) 
sex <- c("male", "female")
antec <-  c("AHT", "DM")
str_glue("This is a {age} years {sex}, diagnosed with {antec}")
```

\

```{r}
data.frame(age, sex, antec) %>% 
  mutate(narrative = str_glue("This is a {age} years {sex}, diagnosed with {antec}"))

```

## Exercise

1. With antecedents below, create a data.frame and produce a proper frequency table. 

    ```{r}
    antec <- c("Arterial hypertension", "arterial hypertension", "ARTERIAL HYPERTENSION",
           "arterial   hypertension ", "arterial hypertension ",  "Pneumonia", 
           "pneumonia", " pneumonia")
    ```

1. With the cip codes below, create a data.frame and compute variables `sex` (with appropriate labels) and `year_of_birth` (should be numeric). Hint: to convert a character vector into a numeric vector, use `as.numeric()`.

    ```{r}
    cip <- c("FEGA0720525000", "ROMA1690913000", "PEJI0910503")
    ```

1. With the DISEHTAE data, compute an additional column having this value for the first patient:  "A 52 years male, diagnosed with AHT on 1998-01-01".  
Hint: in `str_glue()`, factors need to be converted to character, as in:  
`str_glue("This is a {as.character(sex)} patient")`
