---
title: "Dates"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)


```


## Common problems in reading dates

```{r}
d <- rio::import("data/dates.xlsx") 
d
class(d$dob); class(d$v1); class(d$v2)
```


## Defining dates as `Date` ojects

```{r}
library(tidyverse)
dd <- d %>% 
  mutate(dob = as.Date(dob),
         v1 = as.Date(v1, format = "%d/%m/%Y"),           # see ?strptime
         v2 = lubridate::dmy(v2))
dd
sapply(dd, class)
```

## Getting components of `Date` objects

```{r}
library(lubridate)

dd %>% 
  select(dob) %>% 
  mutate(dob_year = year(dob),
         dob_month1 = month(dob),
         dob_month2 = month(dob, label = TRUE),
         dob_day = day(dob),
         dob_wday1 = wday(dob, label = TRUE),
         dob_wday2 = wday(dob, label = TRUE, abbr=FALSE))

```

## Computing intervals

```{r}
ddd <- dd %>% 
  select(-v2) %>% 
  mutate(dif1 = v1 - dob,
         dif2 = as.numeric(v1 - dob),
         age_at_v1 = trunc(as.numeric(v1 - dob)/365.25))
ddd
sapply(ddd, class)
```


## Shifting dates

```{r}
enrollment <- as.Date("2022-01-07")

intervals <- 30 * 0:6
intervals

enrollment + intervals
enrollment + months(0:6)

```

## Shifting dates: non-existing dates

```{r}
enrollment <- as.Date("2022-01-31")

enrollment + months(0:6)
enrollment %m+% months(0:6)



```

## Shifting dates: non-existing dates

```{r}
leap_year(2020)
as.Date("2020-02-29") + years(0:5)
as.Date("2020-02-29") %m+% years(0:5)
```


## Date intervals and overlapping

```{r}
treatment_phase <- interval("2021-01-01", "2021-03-31")
steroids <- interval("2020-12-15", "2020-12-31")
statins <- interval("2021-02-15", today())

int_overlaps(treatment_phase, steroids)
int_overlaps(treatment_phase, statins)
```

## Exercise

1. With the DISEHTAE data:
    + What is the class of `ah_dx_dt` and `data_xtract_dt`?
    + Convert them to to `Date` objects.
    + Compute the years from the AHT diagnostic to the data extraction date.
    + Create a new variable for the week day of the data extraction date.
    
1. Use the `lubridate` function `dmy()` to define a `Date` object from the following character vector. Did it work?  

```{r}
x <- c("05-mar-58", "2/12/57", "28-12-91", "7/jun/93")
```




