---
title: "Summarising rows"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)

```


## Summarising all rows

`summarise(data, varname = function(var), ....)`

```{r}
library(tidyverse)
demo <- rio::import("data/hta.xlsx", sheet = "data") %>% 
  mutate(smoke = factor(smoke, 1:4, c("current", "never", "ex", "unknown")),
         sex = factor(sex, 1:2, c("male", "female"))) %>% 
  select(pid, age, sex, smoke) %>% 
  na.omit()

demo %>% 
  summarise(age_mean = mean(age),
            age_sd   = sd(age),
            cases = n()) 

```


## Summarising groups of rows

`group_by(data, var, ...)`  
`summarise(data, varname = function(var), ....)`

```{r}
demo %>% 
  group_by(smoke) %>% 
  summarise(age_mean = mean(age),
            age_sd   = sd(age),
            cases = n()) 

```


## Grouping rows by more than one variable

```{r}
demo %>% 
  group_by(smoke, sex) %>% 
  summarise(age_mean = mean(age),
            age_sd   = sd(age),
            cases = n()) 

```


## Blood pressure in tidy format

```{r}
bp <- rio::import("data/hta.xlsx", sheet = "data") %>% 
  select(pid, sbp_v1:dbp_v6) %>% 
  pivot_longer(sbp_v1:dbp_v6) %>% 
  separate(name, into = c("var", "visit")) %>% 
  pivot_wider(names_from = var) %>% 
  na.omit() 

head(bp)

```


## Getting fist row in each group

`first(var, ...)`

```{r}
bp %>% 
  arrange(pid, visit) %>% 
  group_by(pid) %>% 
  summarise(sbp_first = first(sbp),
            dbp_first = first(dbp)) %>% 
  head()
```

## Getting last row in each group

`last(var, ...)`

```{r}
bp %>% 
  arrange(pid, visit) %>% 
  group_by(pid) %>% 
  summarise(sbp_last = last(sbp),
            dbp_last = last(dbp)) %>% 
  head()
```


## Getting nth row in each group

`nth(var, n, ...)`

```{r}
bp %>% 
  arrange(pid, visit) %>% 
  group_by(pid) %>% 
  summarise(sbp_3 = nth(sbp, 3),
            dbp_3 = nth(dbp, 3)) %>% 
  head()
```


## Counting values of categorical variables

`count(data, var, ...)`

```{r}
demo %>% count(sex)
demo %>% count(sex, smoke)
```


## Exercises

1. Count the number of men and women for each region.

1. Count the number of diabetic patients of each sex.

1. Get blood pressure data in tidy format (`pid`, `visit`, `sbp` and `dbp`) and discard the first three visits; with the remaining visits, compute for each patient:
    + the mean values of SBP and DBP.
    + the number of rows used to compute each of the means.
    
1. Starting with drug treatments in long format, compute the number of drugs taken by each patient.

1. What is the most frequent number of drugs prescribed to a patient? How many patients had just one drug prescribed? And how many had two, or three? Is there any patient with more than three drugs? 