---
title: "Univariate graphics"
subtitle: "with package `ggformula`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align='center')

```


## Data

```{r}
library(tidyverse)
d <- rio::import("data/hta.xlsx", sheet = "data") %>% 
  mutate(sex = factor(sex, 1:2, c("male", "female")),
         smoke = factor(smoke, 1:4, c("current", "never", "ex-", "unknown")))

```

## Package `ggformula`

Not part of the `tidyverse`

```{r}
library(ggformula)
```

\

Based on `ggplot2` but easier:  

- univariate:  $\qquad$  `gf_plot_type(~ var, data)`  
- multivariate:  $\qquad$  `gf_plot_type(var1 ~ var2, data)`  

\

`data %>% gf_plot_type(...)`  

## Barchart (counts)

```{r }
gf_bar(~ smoke, data = d)
```


## Barchart (counts or percents)

```{r}
library(patchwork)
gf_bar(~ smoke, data = d) + gf_percents(~ smoke, data = d)
```


## Histogram (counts)

```{r}
gf_histogram(~ age, data = d)
```


## Histogram (counts or densities)

```{r}
gf_histogram(~ age, data = d) + gf_dhistogram(~ age, data = d)
```

## Histogram: bins and boundary

```{r}
gf_histogram(~ age, data = d, bins = 10) + 
  gf_histogram(~ age, data = d, bins = 10, boundary = 30)
```


## Density

```{r}
gf_density(~ age, data = d) +
  gf_dens(~ age, data = d)
```


## Boxplots (with color and fill)

```{r fig.height=3}
gf_boxplot( ~ weight, data = d, col = "blue") /
  gf_boxplot( ~ weight, data = d, fill = "lightblue")
```


## Exercise

1. Use `gf_barh()` to produce a barchart of region with blue lines and no fill (hint: `fill = NA`):
    + what is the region that contributes more patients?
    + and what is the region that contributes fewer patients?
1. Same, but with blue fill and no lines.  
1. Look at the distribution of the body mass index:  produce a density plots (top) and boxplots (bottom) in a single graphic output:  
    + is the distribution unimodal? 
    + is it symmetrical? 
    + are there outliers?


