---
title: "Bivariate graphics: group comparisons"
subtitle: "with package `ggformula`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align = 'center',
                      fig.height = 4)

```


## Data

```{r}
library(tidyverse)
library(ggformula)
library(patchwork)

d <- rio::import("data/hta.xlsx", sheet = "data") %>% 
  mutate(sex = factor(sex, 1:2, c("male", "female")),
         smoke = factor(smoke, 1:4, c("current", "never", "ex-", "unknown")),
         bmi = round(weight / height^2, 1))

```


## Jitter plot 

```{r}
d %>% 
  gf_jitter(bmi ~ sex) + coord_flip()
```


## Jitter plot: exclude NA category

```{r fig.height=4}
d %>% 
  filter(!is.na(sex)) %>% 
  gf_jitter(bmi ~ sex)  + coord_flip()
```


## Jitter plot: $\quad$  `width` and `alpha` 

```{r fig.height=4}

set.seed(123)                                       # for reproducible jittering
d %>% 
  filter(!is.na(sex)) %>% 
  gf_jitter(bmi ~ sex, width = .2, alpha = 0.4) + coord_flip()

```


## Boxplots

```{r fig.height=4}
d %>% 
  filter(!is.na(sex)) %>% 
  gf_boxplot(bmi ~ sex, data = d) + coord_flip()
```


## Boxplots: $\quad$  `notch` and `varwidth`

```{r fig.height=4}

d %>% 
  filter(!is.na(sex)) %>% 
  gf_boxplot(bmi ~ sex, notch = TRUE, varwidth = TRUE) + coord_flip()

```

## Violin plot

```{r fig.height=4}

d %>% 
  filter(!is.na(sex)) %>% 
  gf_violin(bmi ~ sex) + coord_flip()

```


## Density plots: using `fill`

```{r fig.height=4}
d %>% 
  filter(!is.na(sex)) %>% 
  gf_density(~ bmi, fill = ~ sex)

```


## Stacked bars

```{r fig.height=4}
dd <- d %>% filter(!is.na(sex) & smoke %in% c("current", "never", "ex-"))
plot1 <- gf_bar( ~ sex , fill = ~ smoke, data = dd)                      # counts 
plot2 <- gf_props(~ sex , fill = ~ smoke, data = dd, position = "fill")  # percentages
plot1 + plot2
```


## Exercise

1. Produce the following plots to compare the distribution of age in men and women:
    + Jitter:  do you see any difference?
    + Boxplot: do you see any difference?
    + Violin: do you see any difference?
    + Density: do you see any difference?

1. Compute age groups with cutpoints `0, 40, 70, Inf`, and produce a stacked barchart to compare age gropups in  men and women. Do you see any difference?
