---
title: "Bivariate graphics: scatterplots and correlograms"
subtitle: "with packages `ggformula` and `corrgram`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align='center')

```


## Data

[Oropharyngeal examination to predict sleep apnea severity](https://pubmed.ncbi.nlm.nih.gov/22006776/)

```{r}
library(tidyverse)
library(ggformula)
library(patchwork)
d <- rio::import("data/OSAS.xls") %>% 
    mutate(gender = factor(gender, 1:0, c("male", "female")),
           ftp = factor(ftp, 1:4, paste("grade", 1:4)),
           tonsils = factor(tonsils, 0:4, paste("grade", 0:4)),
           uvula = factor (uvula, 1:4, c("no hypertrophy", "slight hyp.", 
                                         "moderate hyp.", "marked hyp.")))
head(d, 2)

```


## Scatterplot

```{r fig.width=10, fig.height=4}
a <- gf_point(flow_100_r ~ flow_100_l, data = d, alpha = 0.4)
b <- gf_point(flow_150_r ~ flow_150_l, data = d, alpha = 0.4)
a + b
```


## Excluding the outlier

```{r fig.width=10, fig.height=4}
d <- d %>% filter(flow_150_r < 3000)
a <- gf_point(flow_100_r ~ flow_100_l, data = d, alpha = 0.4)
b <- gf_point(flow_150_r ~ flow_150_l, data = d, alpha = 0.4)
a + b
```



## Scatterplot matrix

```{r fig.width = 7, fig.height= 5}
df <- d %>% select(contains("flow"))                   # data with only flow_* vars
pairs(df)
```


## Scatterplots:  flows (left) vs sides (right)

```{r fig.width = 8, fig.height= 4}
a <- gf_point(flow_150_l ~ flow_100_l, data = d, alpha = 0.4)
b <- gf_point(flow_150_l ~ flow_150_r, data = d, alpha = 0.4)
a + b
```


## Contour plot:  flows (left) vs sides (right)

```{r fig.width = 8, fig.height= 4}
a <- gf_density2d(flow_150_l ~ flow_100_l, data = d, alpha = 0.4)
b <- gf_density2d(flow_150_l ~ flow_150_r, data = d, alpha = 0.4)
a + b
```


## Correlograms

```{r fig.width = 6, fig.height= 5}
library(corrgram)
corrgram(df, lower.panel=panel.ellipse, upper.panel=panel.pts, text.panel=panel.txt) 

```

## Correlograms with Pearson's correlations

```{r fig.width = 6, fig.height= 5}
corrgram(df, lower.panel=panel.cor, upper.panel=panel.shade, text.panel=panel.txt) 
```

- packages corrplot, corrgram, and see https://r-graph-gallery.com/correlogram.html


## Exercise {.smaller}

1. Exclude the outlier and compute the following variables:
    + `flow_100_sum': the sum of flows at 100 Pa in both sides.
    + `flow_150_sum': the sum of flows at 150 Pa in both sides.
    + `flow_100_dif': the difference of flows at 100 Pa (right minus left).
    + `flow_150_dif': the difference of flows at 150 Pa (right minus left).

1. Investigate the following relations:
    + sums at 100 and 150 Pa.
    + differences at 100 and 150 Pa.
    + sum and difference at 100 Pa.
    + sum and difference at 150 Pa.
    
1. Produce a scatterplot matrix with all four vars.

1. Produce a correlogram with all four vars, having points in the lower panel and Pearson's correlations in the upper panel.

