---
title: "Adding dimensions"
subtitle: "color, shape, size and facets"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align='center')

```


## Data

Birth weight data, Baystate Medical Center, Springfield, Mass 1986.

See `?MASS::birthwt`

```{r}
library(tidyverse)
library(ggformula)
library(patchwork)

d <- MASS::birthwt %>% 
  mutate(lwt = round(0.453592 * lwt,1),
         race = factor(race, 1:3, c("white", "black", "other")),
         across(c(low, smoke, ht, ui), factor, 0:1, c("no", "yes")))
head(d, 3)

```


## Multivariate graphics

A graphic has only two axes:

- one for the analysis variable (`Y`),

- another for an explanatory variable (`X`)

\

To add info on other explanatory variables (e.g., potential confounders), use:

- size, color: $\quad \qquad \qquad \qquad \quad$ for continuous variables 

- shape, color, facets: $\qquad \qquad$ for factors

 
## Size and color: $\quad$ continuous variables

```{r fig.width=10, fig.height=4}
a <- gf_jitter(bwt ~ race, data = d, width = 0.2, alpha = 0.4, size = ~lwt)
b <- gf_jitter(bwt ~ race, data = d, width = 0.2, alpha = 0.4, col = ~lwt)
a + b
```


## Color: $\quad$ continuous variables and factors

```{r fig.width=10, fig.height=4}
a <- gf_jitter(bwt ~ race, data = d, width = 0.2, alpha = 0.4, col = ~ age)
b <- gf_jitter(bwt ~ race, data = d, width = 0.2, alpha = 0.4, col = ~ smoke)
a + b

```


## Shape vs color: $\quad$factors

```{r fig.width=10, fig.height=4}
a <- gf_jitter(bwt ~ race, data = d, width = 0.2, alpha = 0.4, col = ~ smoke)
b <- gf_jitter(bwt ~ race, data = d, width = 0.2, alpha = 0.4, shape = ~ smoke)
a + b
```


## Facets:  $\quad$ factors

```{r}
levels(d$smoke) <- c("non-smoker", "smoker")
gf_jitter(bwt ~ race | smoke, data = d, width = 0.2, alpha = 0.4)
```


## Facets:  $\quad$ two factors

```{r}
levels(d$ui) <- c("no UI", "UI")
gf_jitter(bwt ~ race | smoke + ui, data = d, width = 0.2, alpha = 0.4)
```


## Five explanatory vars

```{r}
gf_jitter(bwt ~ race | smoke + ui, data = d, width = 0.2, alpha = 0.4, col = ~ ht, 
          size = ~ lwt)
```


## Exercise {.smaller}

[PrediMed trial](https://www.nejm.org/doi/10.1056/NEJMoa1800389?url_ver=Z39.88-2003&rfr_id=ori:rid:crossref.org&rfr_dat=cr_pub%3dwww.ncbi.nlm.nih.gov)

Men (55 to 80 years of age) or women (60 to 80 years of age) with no cardiovascular disease at enrollment, who had either type 2 diabetes mellitus or at least three of major risk factors.

1. Install and load package `compareGroups` and then run this:
```{r eval = FALSE}
library(compareGroups)
data(predimed)
levels(predimed$group) <- c("Control", "MD+nuts", "MD+VOO") # for shorter levels
```

1. Produce a jitter plot of `age` by `group` using `alpha = 0.2`):  
    + Are all cases within the allowed age range? 
    + Is the distribution of age similar in all three groups?
    + Any feature of the age distribution?

1. Repeat adding info on `sex` using `shape` or `color`. Which do you prefer?

\

(continued in next slide)


## Exercise (cont.){.smaller}

1. Produce a scatterplot of `waist` and `bmi` faceting by `group` and using color for `sex`:
    + Any difference between groups?
    + Any difference between men and women?
    
1. Repeat using `gf_density_2d()`
    + Any difference between men and women?

 
