---
title: "Refining graphics 1"
subtitle: "with package `ggformula`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align='center', fig.height = 4)

```


## Data

```{r}
library(tidyverse)
library(ggformula)

d <- rio::import("data/hta.xlsx", sheet = "data") %>% 
  filter(!is.na(sex)) %>% 
  mutate(sex = factor(sex, 1:2, c("male", "female")),
         smoke = factor(smoke, 1:4, c("current", "never", "ex-", "unknown")), 
         bmi = round(weight / height^2, 1))

```


## Title, subtitle and axis labels

```{r fig.height=4}
plot1 <- gf_jitter(bmi ~ sex, data = d, width = 0.2, alpha = 0.4, color = ~sex,
          title = "Body Mass Index (BMI) by sex",  subtitle = "N = 491",
          ylab = "BMI (Kg/m^2)",  xlab = "")
plot1

```


## Axis limits

```{r fig.height=4}
plot1 %>% 
  gf_lims(y = c(20, 40))

```

## `gf_refine()`

Makes `ggplot2` plot elements chainable with `%>%`.  
Those familiar with `ggplot2` can replace:

```{r eval=FALSE}
a_plot + foo()                          # ggplot2 style: chaining with +
a_plot %>% gf_refine(foo())             # ggformula style: chaining with %>% 
```

\

\

See next slide for an example using `coord_flip()`.   


## Flipping axes:  $\quad$ `coord_flip()`

```{r }

plot1 %>% 
  gf_refine(coord_flip())
  
```


## Overlaying plots

```{r }

plot1 %>% 
  gf_violin(fill = NA) %>% 
  gf_refine(coord_flip())
  
```


## Labelling cases:  $\quad$ `gf_text()` and `gf_label()`

```{r fig.height = 3.8}
plot2 <- gf_point(weight ~ height, data = d, alpha = 0.4, color = ~sex) %>% 
  gf_label(data = filter(d, weight>125), label = ~pid, 
           show.legend=FALSE, nudge_y = 5, nudge_x = - 0.03) %>% 
  gf_text(x = 1.7, y = 145, label = "some outliers here", col = "darkgray")
plot2
```


## Themes

- Control the appearance of (a huge list of) non-data elements of plots.

- There is a number of pre-defined themes such as:  
    + `theme_bw()`  
    + `theme_minimal()`  
    + `theme_classic()`  
    + ... (see `?theme` for the complete list)  
    + ... (additional themes can be found in the `ggthemes` package)  

\

- Pre-defined themes can be applied with `gf_theme()` or `gf_refine()`.

- Themes elements can be adjusted with `gf_theme()`.


## Applying pre-defined themes 

```{r }

plot2 %>% 
  gf_refine(theme_classic())

```


## Adjusting theme elements

```{r}
plot2 %>% 
  gf_refine(theme_classic()) %>% 
  gf_theme(legend.position = "top", legend.title = element_blank())

```





## Exercise

- With the `predimed` data in package `compareGroups`, reproduce this graphic:

    ```{r echo = FALSE}
    library(compareGroups)
    data(predimed)
    set.seed(123)

    p <- gf_jitter(age ~ group , data = predimed, alpha = 0.2, color = ~sex,
                   title = "Age at enrollment by randomized treatment",
                   subtitle = "Predimed trial (n = 6324)",
                   xlab = "",
                   ylab = "Age (years)") %>% 
      gf_violin(fill = NA, col = "darkgray") %>% 
      gf_refine(coord_flip())
    
    p

    ```

- Apply the `theme_minimal()` theme to the previous graphic.


## Exercise (cont.)

- See `?gf_hline` 

- Use this function to draw reference lines at ages 55, 60  and 80:

    ```{r echo = FALSE}
    p %>% 
      gf_theme(theme_minimal()) %>% 
      gf_hline(yintercept = c(55, 60, 80), linetype = 2)

    ```


