---
title: "Refining graphics 2"
subtitle: "with package `forcats` (part of the tidyverse)"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align='center', 
                      fig.width = 10,
                      fig.height = 4)

```


## Data

```{r}
library(tidyverse)
library(ggformula)
library(patchwork)

d <- rio::import("data/hta.xlsx", sheet = "data") %>% 
  select(age, sex, region) %>% 
  mutate(sex = factor(sex, 1:2, c("male", "female")),
         region = factor(region, 
                  levels = 1:17, 
                  labels = c("Andalucía", "Aragón", "Asturias",
                             "Baleares", "Canarias", "Cantabria",
                             "Castilla-La Mancha", "Castilla-León",
                             "Catalunya", "Extremadura", "Galicia", 
                             "La Rioja", "Madrid", "Murcia", "Navarra",
                             "País Vasco", "Valencia"))) %>% 
  na.omit()

```


## Factors: ordering levels by frequency

```{r}
gf_barh(~ region, data = d) +
gf_barh(~ fct_infreq(region), data = d)
```


## Factors: reversing order of levels

```{r}
gf_barh(~ fct_infreq(region), data = d) +
gf_barh(~ fct_rev(fct_infreq(region)), data = d, ylab = "Region")
```


## Factors: reversing order of levels

```{r fig.height = 3.5}

bw <- MASS::birthwt %>% 
  mutate(low = factor(low, 0:1, c("no", "yes")),
         race = factor(race, 1:3, c("white", "black", "other")))
a <- gf_jitter(age ~ race, color = ~ low, data = bw, width = .2) + coord_flip()
b <- gf_jitter(age ~ race, color = ~ fct_rev(low), data = bw,  width = .2) + coord_flip()
a + b
```


## Factors: ordering by other variables

```{r fig.height=3.5}
mar <- d %>% 
  group_by(region) %>% 
  summarize(mean_age = mean(age))        # mean age by region
gf_point(region ~ mean_age, data = mar) +
gf_point(fct_reorder(region, mean_age) ~ mean_age, data = mar, ylab = "")

```


## Factors: recoding levels

```{r fig.height=3}
d %>% 
  mutate(region = fct_recode(region,
                             "Principado de Asturias" = "Asturias",
                             "Comunidad Foral de Navarra" = "Navarra",
                             "Comunitat Valenciana" = "Valencia", 
                             "Castilla y León" = "Castilla-León",
                             "Cominudad de Madrid" = "Madrid")) %>% 
  gf_barh(~fct_infreq(region) %>% fct_rev(), ylab = "Region")
```



## Factors: collapsing levels

```{r fig.height=2.5}
d %>% 
  mutate(region = fct_collapse(region,
                               "North" = c("Galicia", "Asturias", "Cantabria", "País Vasco",
                                           "Navarra", "La Rioja", "Aragón"),
                               "East" = c("Catalunya", "Valencia", "Murcia"),
                               "South" =  "Andalucía", 
                               "Middle-West" = c("Castilla-La Mancha", "Castilla-León", 
                                                 "Madrid", "Extremadura"))) %>% 
  gf_barh(~ fct_infreq(region) %>% fct_rev(), ylab = "Region")
```


## Factors with sparse levels 

```{r}
d <- data.frame(pid = 1:146,
                center = factor(paste("Hospital",
                                      rep(LETTERS[1:8], 
                                          times = c(51, 26, 4, 5, 6, 32, 16, 6)))))
d %>% count(center)
```


## Factors: collapsing sparse levels 

```{r}
gf_barh(~ center %>% fct_rev(), data = d, ylab = "") +
gf_barh(~ fct_lump_min(center, 10) %>% fct_rev(), data = d, ylab = "") +
gf_barh(~ fct_lump_min(center, 10, other_level = "Small centers") %>% fct_rev(), 
        data = d, ylab = "") 
```


## Exercise 

- With the `birthwt` data, produce these graphic:

```{r echo = FALSE}

bw <- MASS::birthwt %>% 
  mutate(smoke = factor(smoke, 0:1, c("non-smoker", "smoker")),
         race = factor(race, 1:3, c("white", "black", "other")))

gf_jitter(bwt ~ race, data = bw, width = 0.2, alpha = 0.4, 
          col = ~fct_rev(smoke),
          ylab = "Birth weight (g)") %>% 
  gf_theme(legend.title = element_blank())

```

## Exercise (cont.)

- The following code reads data on the dosages of a drug given to patients in a study, and redefines variable `dose` so as to contain numbers only. With this data, produce the barcharts shown below (low: 25, 50; medium: 100, high: 150, 200)

```{r }
d <- rio::import("data/doses.xlsx") %>% 
  mutate(dose = factor(str_extract(dose, "[:digit:]+"),
                       levels = c("25", "50", "100", "150", "200")))
```

```{r echo=FALSE, fig.height=3}
gf_barh(~ dose, data = d) +

d %>% 
  mutate(dose_range = fct_collapse(dose,
                                   "low" = c("25", "50"),
                                   "medium" = "100",
                                   "high" = c("150", "200"))) %>% 
  gf_barh(~ dose_range,
          ylab = "dose range")

```




