---
title: "Line plots (for lab data)"
subtitle: "with package `ggformula`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align='center', 
                      fig.width = 10,
                      fig.height = 4)

```


## Data

```{r}
library(tidyverse)
library(ggformula)

d <- rio::import("data/labs.xlsx")

head(d)

```


## Patient profiles

```{r}

crea <- d %>% filter(var == "creatinina" & !is.na(value)) 

gf_line(value ~ visit | treatment, group = ~ subject_id, 
        data = crea, ylab = "creatinina (mg/dl)") 

```

## Adding reference line with `gf_hline()`

```{r}
gf_line(value ~ visit | treatment, group = ~ subject_id, 
        data = crea, ylab = "creatinina (mg/dl)") %>% 
  gf_hline(yintercept = 1.8, lty=3, lwd=0.5) 
```


## Plotting by site with `gf_facet_grid()`

```{r}

gf_line(value ~ visit | treatment, group = ~ subject_id, 
        data = crea, ylab = "creatinina (mg/dl)") %>% 
  gf_hline(yintercept = 1.8, lty=3, lwd=0.5) %>% 
  gf_facet_grid(site ~ treatment)


```


## Identify patients with elevations

```{r}

pwe <- crea %>% filter(value > 1.8) %>% pull(subject_id) %>% unique()

crea <-  crea %>% 
  mutate(elevated = ifelse(subject_id %in% pwe, "yes", "no")) 

head(crea)


```

## Using color for cases with elevations

```{r}

gf_line(value ~ visit | treatment, group = ~ subject_id, col = ~ elevated,
        data = crea, ylab = "creatinina (mg/dl)") %>% 
  gf_hline(yintercept = 1.8, lty=3, lwd=0.5) %>% 
  gf_facet_grid(site ~ treatment) %>% gf_theme(legend.position = "none")

```


## Better assignment of color

```{r}

crea_plot <- gf_line(value ~ visit | treatment, group = ~ subject_id, col = ~ fct_rev(elevated),
                     data = crea, ylab = "creatinina (mg/dl)") %>% 
  gf_hline(yintercept = 1.8, lty=3, lwd=0.5) %>% 
  gf_facet_grid(site ~ treatment) %>% gf_theme(legend.position = "none")
crea_plot

```


## Labeling cases with elevations

```{r}

crea_plot %>% 
  gf_label(data = filter(crea, value > 1.8), label = ~ subject_id, 
           show.legend = FALSE, nudge_y = 1)

```


## Identifying last visits of cases with elevations

```{r}

 crea %>% 
    filter(elevated == "yes") %>% 
    group_by(subject_id) %>% 
    summarise(visit = last(visit)) %>% 
    ungroup() %>% 
    left_join(crea) -> last_visits

last_visits

```


## Labeling last visits only

```{r}
crea_plot %>% 
  gf_text(data = last_visits, label = ~ subject_id, 
           show.legend = FALSE, nudge_x = 0.2)
```


## Exercise

1. Produce a similar plot for potassium (reference range from 2.5 to 5.5 mEq/l).

1. Repeat using `gf_label()` instead of `gf_text()`. 

1. Download package `ggrepel` and change the entire `gf_label()` line by this one:

```{r eval = FALSE}
gf_refine(ggrepel::geom_label_repel(data = last_visits, 
                                    aes(label = subject_id)))
```

What is the difference?