---
title: "Other plots"
subtitle: "(several packages)"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align='center', 
                      fig.width = 10,
                      fig.height = 4)

```


## Data

[Coronavirus: the first three months as it happened. Nature 2020 Apr](https://www.nature.com/articles/d41586-020-00154-w)

```{r}
library(tidyverse)

d <- rio::import("data/covid_symptoms.xlsx") %>% 
  select(-symptoms) %>% 
  uncount(count) %>% 
  mutate(case = row_number()) %>% 
  select(case, everything())

head(d, 5)

```


## Venn diagrams (dichotomous vars)

```{r}

library(ggvenn)
ggvenn(list(Fatigue = d %>% filter(Fatigue) %>% pull(case), 
          Cough = d %>% filter(Cough) %>% pull(case),
          Fever = d %>% filter(Fever) %>% pull(case)), 
  fill_color = c("darkorange", "darkcyan", "darkgreen"), stroke_size = 0.5, set_name_size = 5)

```


## UpSet plot (signs & symptoms)

```{r}
library(ComplexUpset)
ds <- d %>% select(-case) 
upset(data = ds, intersect = names(ds), width_ratio = 0.125, name = "Symptoms") +
  labs(title = "Co-Occurence of COVID-19 Symptoms (n = 1764)")
```


## Mosaic plots (categorical vars)

```{r}
library(vcd)
mosaic(Improved ~ Treatment | Sex, data = Arthritis, 
       zero_size = 0, highlighting_direction = "right")
```


## Likert scales (ordinal vars) 

```{r}
library(likert); data(pisaitems); 
d <-  pisaitems %>% select(contains("ST24Q")) 
names(d) <- c("I read only if I have to", 
              "Reading is one of my favourite hobbies", 
              "I like talking about books with other people", 
              "I find it hard to fard to finish books", 
              "I feel happy if I receive a book as present", 
              "For me reading is a waste of time", 
              "I enjoy going to a bookstore or a library", 
              "I read only to get information I need",
              "I cannot sit still and read for more than a few minutes", 
              "I like to express my opinions about books I have read", 
              "I like to exchange books with my friends")
```

Items on reading attitudes (strongly disagree, disagree, agree, strongly agree)


## Likert scales (ordinal vars)

```{r}

likert(d) %>% 
  plot(legend.position = "right")

```

## QQ plot (quantitative var, assess normality)

```{r}
library(tidyverse); library(ggformula); library(patchwork)
gf_qq(~bwt, data = MASS::birthwt, title = "Birth weight (g)") %>% gf_qqline() +
gf_density(~bwt, data = MASS::birthwt) /
gf_boxplot(~bwt, data = MASS::birthwt)
```

## QQ plot (quantitative var, assess normality)

```{r}
gf_qq(~lwt, data = MASS::birthwt, title = "Mother's weight (pounds)") %>% gf_qqline() +
gf_density(~lwt, data = MASS::birthwt) /
gf_boxplot(~lwt, data = MASS::birthwt)
```



## Exercise {.smaller}

1. Run the following code to get the Titatic data in a dataframe where each row is a person:

    ```{r}
    d <- Titanic %>% as_tibble() %>% tidyr::uncount(n)
    ```

    Then produce mosaic plots, and interpret the result, to explore survival by ...

    - Sex

    - Age

    - Class

    - Sex, given class

    - Age, given class


1. Use QQ plots to assess approximate normality of time to event and BMI in the `predimed` dataset.

```{r}
library(compareGroups)
data(predimed)
```

