---
title: "Refining graphics 3"
subtitle: "with package `ggformula`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA,
                      fig.align='center', 
                      fig.width = 10,
                      fig.height = 4)

```


## Data

```{r}
library(tidyverse)
library(ggformula)
library(patchwork)

d <- rio::import("data/labs.xlsx") %>% 
tidyr::pivot_wider(values_from = "value", names_from = "var")

head(d, 4)
```


## Transformations


```{r}
gf_density(~ GGT, data = d, fill = ~treatment) +
gf_boxplot(GGT ~ treatment, data = d, fill = ~treatment) 

```

## Transformations of coordinates

```{r}
gf_boxplot(GGT ~ treatment, data = d, fill = ~treatment) +
gf_boxplot(GGT ~ treatment, data = d, fill = ~treatment) %>% 
  gf_refine(coord_trans(y = "log10"))

```


## Transformations of values

```{r}
gf_boxplot(GGT ~ treatment, data = d, fill = ~treatment) %>% 
  gf_refine(scale_y_log10()) +
gf_boxplot(log10(GGT) ~ treatment, data = d, fill = ~treatment)

```


## Colors

```{r fig.height = 3}
p <- gf_boxplot(GGT ~ treatment, data = d, color = ~treatment) 
p %>% gf_refine(scale_color_manual(values = c("orange", "darkblue"))) +
p %>% gf_refine(scale_color_manual(values = c("#FFA500", "#00008B")))
```

- See available color names with `colors()`



## Color palettes  $\quad$ (`RColorBrewer` and more)

```{r fig.height = 2.5}
library(RColorBrewer)
p %>% gf_refine(scale_color_brewer(palette="Dark2"),
                coord_flip()) 

```

- See all RColorBrewer palettes with `RColorBrewer::display.brewer.all()`

- Many other palettes available through package
[`paletteer`](https://emilhvitfeldt.github.io/paletteer/) 



## More on graphics

`ggformula`: $\quad$ RStudio tutorials and [this article](http://www.mosaic-web.org/ggformula/articles/pkgdown/ggformula-long.html)

`ggplot2`:

- [A ggplot2 tutorial for beautiful plotting in R](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/), by Cédric Sherer.

- [Introduction to data visualization with ggplot2](https://www.datacamp.com/courses/introduction-to-data-visualization-with-ggplot2), Datacamp online course.

- [The R graph gallery](https://r-graph-gallery.com/), by Yan Holtz.

\

`ggpubr`: $\quad$ [Publication Ready Plots](http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/).

\

Other plotting systems in R: $\quad$ [Base R](http://www.sthda.com/english/wiki/r-base-graphs) and [`lattice`](http://www.sthda.com/english/wiki/lattice-graphs).


## Exercise {.smaller}

- Run this code to read the `OSAS.xls` data and compute `ccg`, a grouped version of the cervical circumference.  In the R graph gallery, look for the type of graphic shown below. Copy the code of the __Most basic ridgeline__, and adapt it to produce the graphic shown below.

```{r}
d <- rio::import("data/OSAS.xls")  %>% 
  mutate(ccg = cut(cervical_circumference, breaks = c(30, 38.5, 41.5, 44, 55)))
```


```{r echo = FALSE}
library(ggridges)
library(ggplot2)

d %>% 
  ggplot(aes(x = ahi, y = ccg, fill = ccg)) +
  geom_density_ridges() +
  theme(legend.position = "none")
```


## Exercise (cont.) {.smaller}

- Install package `ggpubr`. Then look at the first example in [this tutorial](http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/76-add-p-values-and-significance-levels-to-ggplots/#compare-two-independent-groups), and try to produce the graphics below. 

```{r echo = FALSE}
library(ggpubr)
d %>% 
  mutate(gender = factor(gender, 0:1, c("female", "male"))) %>% 
  gf_boxplot(ahi ~ gender, fill = ~gender) %>% 
  gf_refine(stat_compare_means()) +

d %>% 
  gf_boxplot(ahi ~ ccg, fill = ~ccg) %>% 
  gf_refine(stat_compare_means())
```


## Review exercise 

- Produce this graphic with the `aht.xls` data.  
Hint: use `show_col(hue_pal()(2))` to see how to specify the colors for the reference lines.

```{r echo = FALSE}
d <- rio::import("data/hta.xlsx") %>% 
  select(pid, sbp_v1:dbp_v6) %>%
  pivot_longer(sbp_v1:dbp_v6, names_to = "var", values_to = "bp")  %>% 
  na.omit() %>% 
  separate(var, into = c("variable", "visit")) 

library(ggformula)
gf_boxplot(bp ~ visit, fill = ~fct_rev(variable), data = d) %>% 
  gf_hline(yintercept = 140, lty = 2, col = "#F8766D") %>% 
  gf_hline(yintercept = 90, lty = 2, col = "#00BFC4") %>% 
  gf_labs(y = "BP (mmHg") %>% 
  gf_theme(legend.title = element_blank())
```


