---
title: "Measuring association"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning=FALSE,
                      comment=NA)

library(dplyr)
library(ggformula)
library(mosaic)
```

## Association = Probabilistic relationship

Scenario     Formally              Implication (meaning)
--------     ---------------       -----------------------
Independent  Pr(Y|X) = Pr(Y)       Distibution of Y is equal for all X
Related      Pr(Y|X) $\ne$ Pr(Y)       Distribution of Y is NOT equal for all X

</br>

Cases:

- X and Y are both categorical
- X and Y are both quantitative
- One of each type 

## Independence

Distibution of Y is equal for all X

```{r fig.height=3.5, fig.width=10, fig.align='center'}

par(mar=c(4,4,2,2))


# Independence ----

# Categorical
X <- rep(LETTERS[1:3], 100)
Y <- c(rep("A", 75), rep("B", 225))

p1 <- gf_props(~ X, fill= ~ Y, position="fill") +
  theme(legend.position="none") +
  ggtitle("Categorical X and Y")

# One of each type
set.seed(1)
Y2 <- 100 +  rnorm(100)
Y2 <- rep(Y2,3)
X2 <- rep(LETTERS[1:3], 100)

p2 <- gf_boxplot(Y2 ~ X2) +
  xlab("X") + ylab("Y") +
    theme(axis.text.y=element_blank(),axis.ticks.y=element_blank()) +
  ggtitle("Categorical X and quantitative Y")


# Quantis 
set.seed(123)
X3 <- seq(0.25,100,0.25)
Y3 <- 100 +  rnorm(length(X3))
p3 <- gf_point(Y3 ~ X3) +
  xlab("X") + ylab("Y") +
  theme(axis.text.x=element_blank(), axis.text.y=element_blank(),
        axis.ticks.x=element_blank(), axis.ticks.y=element_blank()) +
  ggtitle("Quantitative X and Y")


library(gridExtra)
grid.arrange(p1, p2, p3, ncol = 3)

```

Any other pattern is non-independence (association)

## Association measures/analysis

```{r}
tibble(Variables = c("Quantitative (both)", "Categorical (both)", "One quantitative and one categorical"),
          `Measures/analysis` = c("covariance, correlation coefficients", "chi-square, RD, RR, OR, NNT", "mean difference")
          ) %>%
  knitr::kable(format = "markdown")
```


## Categorical X and Y 

<div class="columns-2">

```{r fig.height=3.5, fig.width=4}
d <- read.table("https://raw.githubusercontent.com/acobos/Datasets/master/Genotype_and%20Clozapine.txt", 
                header = TRUE)

gf_props(~ Genotype, fill= ~ Response, data = d, position="fill") 

```

</br> </br> </br> </br> 

Contingency table

```{r}
tally(Response ~ Genotype, data=d)

```

Marginals (row and column totals)

```{r}


addmargins(tally(Response ~ Genotype, data=d))

```


## Expected in case of independence

```{r}
ct <- tally(Response ~ Genotype, data=d)
ctm <- addmargins(ct)
dimnames(ctm)[[1]][3] <- "total"
dimnames(ctm)[[2]][4] <- "total"

qz <- ctm
qz[1:2,1:3] <- rep(NA,6)
qz
```

</br>

- Overall probability of `Response = 178/270 =` ``r round(178/270,3)``
- Apply it to each Genotype:
    - AA: `45 x 178/270 = 29.7`
    - AG: `125 x 178/270 = 82.4`
    - GG: `100 x 178/270 = 65.9`
- Similarly for `Non-response`
    
## Pearson's chi-square ($\chi^2$)

<div class="columns-2">

Observed (O)

```{r}
ct
```

Expected (E)

```{r}
round(chisq.test(ct)$expected,1)
```

$\chi^2 = \sum\limits_{i=1}^{r \times c} \frac{(O_i - E_i)^2}{E_i} = 12.4$

_Distance_ from independence:

- Independence:  $\chi^2=0$
- Association:  $\chi^2>0$ 
    - the higher the $\chi^2$ value, the stronger the association  
    - but it also depends on the *degrees of freedom*: $(r-1) \times (c-1)$

<center>
$r$ rows and $c$ columns 
</center>



## Tables 2 $\times$ 2: $\quad$ RD, RR, and NNT

```{r 2by2, echo=FALSE}
# t <- as.table(matrix(c(75,50,25,50), nrow=2, ncol=2,
#                      dimnames=list(c("Experimental (E)","Control (C)"),
#                                    c("Success (S)", "Failure (F)"))))
# 
# t(t)

data.frame(A = c("75", "25", "100", " ", ".75"),
           B = c("50", "50", "100", " ", ".50")) -> x

row.names(x) <- c("Success (S)", "Failure (F)", "Total", " ", "P(S)")
names(x) <- paste0("group_", names(x))
x

```

Derived from **proportions** of **success** $P_S$: 

- Risk difference (RD): $\quad P_{S|A} - P_{S|B} = 0.75 - 0.50 = 0.25$

- Relative risk (RR): $\quad \frac{P_{S|A}}{P_{S|B}} = \frac{0.75}{0.50} = 1.5$  

- Number needed to treat (NNT):  $\quad \frac{1}{\text{RD}} = \frac{1}{0.25} = 4$



## Tables 2 $\times$ 2:   $\quad$ OR

```{r}

x <- bind_rows(x,
               data.frame(group_A = "3", group_B = "1"))
row.names(x)[6] <- "O(S)"
x
```


Derived from **odds** of **success** $O_S$:  

- Odds ratio (OR): $\frac{O_{S|A}}{O_{S|B}} = \frac{3}{1} = 3$  

Note this equals cross product:

$\frac{O_{S|A}}{O_{S|B}} = \frac{75/25}{50/50} = \frac{75 \times 50}{25 \times 50} = 3$


## Categorical X and quantitative Y

<div class="columns-2">

Weight at birth (grams) and smoking during pregnancy

```{r fig.height=4.5, fig.width=5}
d <- MASS::birthwt
d$smoke <- factor(d$smoke, levels = c(1:0), labels = c("Yes", "No"))

d %>%
    group_by(smoke) %>%
    summarise(bwt = round(mean(bwt))) -> means

library(ggformula)
gf_boxplot(bwt ~ smoke, data=d, col = ~ smoke) %>%
  gf_point(data = means, color = ~ smoke, shape = 15) 
# %>%
#   gf_refine(coord_flip())

library(mosaic)

cat("Means per group:")
as.data.frame(means)


round(diffmean(bwt ~ smoke, data=d)) -> dif
as.matrix(means) -> k
cat("Mean difference:", k[2,2] , "-", k[1,2], "=", dif, "grams")
```

## Quantitative X and Y:  Covariance

Differences wrt the mean of each variable

```{r covar, warning=FALSE, message=FALSE, fig.align='center', fig.height=4, fig.width=6}

library(readr)
d <- read_table2("https://raw.githubusercontent.com/acobos/Datasets/master/Anthropometric_measures.txt", 
    skip = 15)
par(mar=c(2,2,2,2), mfrow=c(1,1))
with (d,plot(Waist,Thorax,las=1, col="gray", xlab="x", ylab="y", cex=.7, xaxt='n', yaxt='n'))
abline(v=mean(d$Waist),col="darkgreen",lty=2)
abline(h=mean(d$Thorax),col="darkred",lty=2)
x = 113.9;  y = 121.6
points(x,y, pch = 19)
arrows(mean(d$Waist),y,x,y, lty=2, length = 0, col="gray")
text(mean(d$Waist), y+5, labels=expression(x-bar(x)), pos=4,col="darkgreen")
arrows(x,mean(d$Thorax),x,y, lty=2, length = 0, col="gray")
text(115, 112, labels=expression(y-bar(y)), pos=4, col="darkred")

text(140,85,"-",cex=2, col="darkblue")
text(140,130,"+",cex=2, col="darkblue")
text(80,85,"+",cex=2, col="darkblue")
text(80,130,"-",cex=2, col="darkblue")
mtext(expression(bar(x)), side=3, at= mean(d$Waist), col="darkgreen")
mtext(expression(bar(y)), side=4, at= mean(d$Thorax), col="darkred", las=1)

```

Covariance(X,Y) = $\frac{1}{n} \sum{(x-\bar{x}) \times (y-\bar{y})}$

## Covariance 

```{r covpatterns, warning=FALSE, message=FALSE, fig.align='center', fig.height=5, fig.width=9}

par(mfrow=c(2,3), mar=c(2,2,3,2))

x = rnorm(1000)

sp = function(slope,error,tit,text) {
y = slope*x +  error * rnorm(1000)
prod = (x-mean(x)) * (y-mean(y))
color = ifelse(prod>0,"navy","red")

plot(x,y, 
     xaxt="n", yaxt="n", xlab="",ylab="", 
     xlim=c(-3,3), ylim=c(-3,3), pch=20, col=color)
title(main=tit, cex.main=.9)
abline(h=mean(y), v=mean(x), col="darkgray")
mtext(text, side=3, cex=.7) 
}

sp(1,0.5,"Relaci?n directa: COV >>> 0","Gran predominio +")

sp(-1,0.5,"Relaci?n inversa: COV <<< 0","Gran predominio -")

sp(0,0.5,"Independencia: COV aprox. 0","No predomina ning?n signo")


sp(1,1.5,"Relaci?n directa: COV > 0","Ligero predominio +")

sp(-1,1.5,"Relaci?n inversa: COV < 0","Ligero predominio -")

sp(0,1.5,"Independencia: COV aprox. 0","No predomina ning?n signo")
```



## Correlation coefficients

Measure *linear* relation (not *any* relation!):

- Pearson's $r$:  $\quad \frac{Cov(X,Y)}{S_X  S_Y}$

- Spearman's $r_s$: $\quad$ Pearson's $r$ of the *ranks* of X and Y

For both, range is  -1 to +1

Care with non-linearities and outliers, ALWAYS inspect scatterplot:

- linear with no outliers:  Pearsons $r$ 

- outliers, or non-linear but *monotonic*: Spearman's $r_s$

- non-linear, non-monotonic: neither!

## Patterns of relationship

```{r fig.align='center', fig.height=5}


my_plot <- function (X, Y) {
plot(X, Y, xaxt='n', yaxt='n')
r <- paste("r =", round(cor(X,Y),2))
rs <- paste("r_s =" , round(cor(X,Y, method="spearman"),2))
title(paste(r, rs, sep = ", "))
}

par(mfrow=c(2,3), mar=c(2,2,3,2))

my_plot(d$Waist, d$Thorax)

my_plot(d$Wrist, d$Biceps)

my_plot(d$Age, d$Weight)

my_plot(d$Ankle, d$Knee)

y <- exp(x)
my_plot(x, y)

x <- seq(-10, 10)
y <- -x^2
my_plot(x, y)


```

