---
title: "Categorical variables"
subtitle: "Independent samples"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)
```

## Birth weights

```{r}
library(tidyverse)

d <- MASS::birthwt %>% 
  mutate(race = factor(race, levels = 1:3, labels = c("white", "black", "other")),
         low = factor(low, levels = c(1,0), labels = c("Low", "Normal")),
         smoke = factor(smoke, levels = c(1,0), labels = c("smoker", "non_smoker")))

head(d)        

```

## Birth weight related to smoking?

```{r}
library(mosaic)
x <- tally(low ~ smoke, data = d)    # save the contingency table for later use
x
tally(low ~ smoke, data = d, format="percent")
```

## Chi-square test (and continuity correction)

- Use function `chisq.test()` on the contingency table of counts

```{r}
chisq.test(x, correct=FALSE)   # x is the contingency table
chisq.test(x)  
```

## Expected frequencies

- Chi-square test is not safe if *expected frequencies* are too low, e.g.:  
    - very low in some cell (e.g. < 3), or
    - low in several cells (e.g. < 5)
    
```{r}
res <- chisq.test(x, correct=FALSE)  # save test result (which is a list)
names(res)                           # see names of list elements
res$expected                         # get list element "expected" 
```

## Expected frequencies: warning when too low


```{r warning=TRUE}
y <- tally(low ~ ht, data = d); y   # now look at hypertension (ht)
chisq.test(y)
```

## Fisher's test

- better when expected frequencies are too low

```{r}
fisher.test(y)
```


## How strong is the association?

You MUST provide some measure of association measure, and 95% CI

Choose one, depending on the study *design*:

Design           Measures
---------------  ---------------------------------------------------------------------------------------
Clinical trial   Risk difference (RD), Relative risk (RR), Odds ratio (OR), Number needed to treat (NNT)
Cross-sectional  Risk difference (RD), Relative risk (RR), Odds ratio (OR)
Cohort           Risk difference (RD), Relative risk (RR), Odds ratio (OR)
Case-control     Odds ratio (OR)
---------------  ---------------------------------------------------------------------------------------
    
## Risk difference (RD)

```{r}
res <- prop.test(low ~ smoke, data = d, correct=FALSE) 
confint(res)

# look at the proportions table to interpret prop1, prop2, and 
# the CI of the difference (lower, upper)
tally(low ~ smoke, data = d, format="proportion", margins = TRUE)

```

## Relative risk (RR) and Odds ratio (OR)

Functions `riskratio()` and `oddsratio()` in package `epitools`:  
- They compare ROWS of a contingency table  
- If you want to compare COLUMNS, first transpose the table with `t()`  

```{r}

x
t(x)   # transpose function

```

## Relative risk (RR) 

Proportions of 2nd column (row 2 vs row 1): $\quad$ RR = (86/115) / (44/74) = 1.26

```{r}
library(epitools)
riskratio(t(x))   
```

## Relative risk (RR) 

Option `rev="columns"` reverses columns:  $\quad$ RR = (29/115) / (30/74) = 0.62

```{r}
riskratio(t(x), rev="columns")     
```

## Relative risk (RR) 

Option `rev="both"` reverses rows & columns:  $\quad$ RR = (30/74) / (29/115) = 1.61

```{r}
riskratio(t(x), rev="both")     
```

## Odds ratio (OR)

Same options for odds:  $\quad$ OR = (30/44) / (29/86) = 2.02

```{r}
oddsratio(t(x), rev="both", method = "wald")    # note method for usual crossproduct 
```


## Exercise

Is low weight at birth related to race? 

1. Answer the question based on the result of a test

2. Compute the RR (95% CI) for the probability of low weight in:

    - black vs white
    - other vs white
    

3. What if you merge levels `black` and `other` into a single category?  
Use the following code and repeat the analysis with `race_2`:

```{r }
d <- mutate(d, race_2 = ifelse(d$race == "white", "white", "black/other"))
```




