---
title: "User defined functions"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)
```

## User defined functions

- R has lots of predefined functions, but you can define new functions as well

- Define a function using function ```function()```, and assigning the result to a *name* of our choice 


```{r eval=FALSE}
my_function <- function() {              # function definition
                                         #    here the R code doing something
}                                        # end of function definition

```

- You will later call the function using its name

```{r eval=FALSE}
my_function()                            # function call

```



## A simple function

We can define one (or more) argument(s) within the brackets

```{r}
my_function <- function(x) {        
  c("mean" = mean(x),  "SD" = sd(x),  "Median" = median(x),  "IQR" = IQR(x))
}                                  

vector1 <- 1:10;  vector2 <- 11:20
my_function(vector1) 
my_function(vector2) 
```

## Improving the output: rounding

```{r}
my_function <- function(x) {        
  res <- c("mean" = mean(x),  "SD" = sd(x),  "Median" = median(x),  "IQR" = IQR(x))
  round(res, 2)
}                                  

my_function(vector1) 
```

</br>

More flexibility: 

Could we allow for a different number of decimals? 

## Arguments with default value

```{r}
my_function <- function(x, decimals=2) {        
  res <- c("mean" = mean(x),  "SD" = sd(x),  "Median" = median(x),  "IQR" = IQR(x))
  round(res, decimals)
}                                  

my_function(vector1) 
my_function(vector1, 4) 
 
```


## What the function returns

A function always returns the result of its *last line* of code

Why this does not produce the desired result?

```{r}
my_function <- function(x, decimals=2) { 
  mean(x)
  sd(x)           # last line of the function body will be returned
}     

my_function(1:5)
```

The mean was actually computed, but not returned !

## Births, again

```{r}
library(mosaic)
data(Births78)
library(dplyr)
d <- Births78 %>%
  mutate(day = recode(day_of_week, 
                      `1` = "Sun",
                      `7` = "Sat",
                      .default = "Working"),
         day= factor(day, levels = c("Working", "Sat", "Sun"))) %>%
  select( -day_of_week) #, -day_of_year)
```


## A first model and its graphical representation

```{r fig.align='center', fig.height=3.5}
model <- lm(births ~ day, data = d)                  # fit the model
d$pred <- predict(model)                            # save predictions
gf_point(births ~ date, col = ~ day, data = d) %>%    # plot the data
  gf_line(pred ~ date)                              # with predictions

```

## What will you do for a different model?

1. Copy paste
2. Change what is needed: just the model formula

```{r eval=FALSE}
model <- lm(births ~ day, data = d)                  
d$pred <- predict(model)                            
gf_point(births ~ date, col = ~ day, data = d) %>%    
  gf_line(pred ~ date)                              

model <- lm(births ~ day + month, data = d)         # change model formula
d$pred <- predict(model)                            
gf_point(births ~ date, col = ~ day, data = d) %>%    
  gf_line(pred ~ date)                         
```

It's ok, but if you need to fit other models, there is a better solution:

write a function

## Writing functions

Methodology: 

1. Solve a particular case (as we did)

2. Identify the part to change, if copy paste (as we did)

3. Give a name to the part to change, replace, and test

4. Wrap the code inside a function, and use this name as argument

## Step 3. Name the part to change, and replace

```{r fig.align='center', fig.height=3.5}
the_formula <- births ~ day + month         # give a name for the formula

model <- lm(the_formula, data = d)          # replace
d$pred <- predict(model)                            
gf_point(births ~ date, col = ~ day, data = d) %>%    
  gf_line(pred ~ date)                         
```

## Step 4. Wrap inside a function, with argument

```{r}
my_function <- function(the_formula) {      # start function definition (argument_name)
  
  model <- lm(the_formula, data = d)          
  d$pred <- predict(model)                            
  gf_point(births ~ date, col = ~ day, data = d) %>%    
    gf_line(pred ~ date)                            
  
}                                        # function end


```

## Call the function...

```{r fig.align='center'}
my_function(births ~ day + month)
```

## ...as many times as you need

```{r fig.align='center'}
my_function(births ~ day + poly(day_of_year,5))
```

