---
title: "More Rmarkdown"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)

# packages
library(dplyr)
library(mosaic)
library(ggformula)

d <- MASS::birthwt %>%
  mutate(low = factor(low, 
                     levels = 1:0, 
                     labels = c("low", "normal")),
         race = factor(race, 
                       levels = 1:3,
                       labels = c("white", "black", "other")),
         smoke = factor(smoke, 
                        levels = 1:0, 
                        labels = c("smoker", "non-smoker")),
         ui = factor(ui, 
                      levels = 1:0, 
                      labels = c("UI", "no-UI")),
         ht = factor(ht, 
                     levels = 1:0, 
                     labels = c("AHT", "no-AHT"))) %>% 
  rename(bw = low)

```

## Tables

Many packages with functions to build tables with R:

- `xtable::xtable()` (useful for pdf outputs)

- `knitr::kable()` and `kableExtra` (for HTML output)

- `compareGroups::descrTable()` (for both pdf and HTML outputs)

- ...and many more.

## compareGroups::descrTable() {.smaller}

```{r table1}
library(compareGroups)
descrTable(smoke ~ bwt + bw + race + ht + ui, data = d,
           show.p.overall = FALSE,
           hide.no = c("no-AHT", "no-UI")) %>%   
  export2md(caption = "Descriptive analysis by smoking status") 

```

## Defining labels for variables

With function `label()` from package `Hmisc`

```{r}
Hmisc::label(d$bwt) <- "Birth weight (grams)"
Hmisc::label(d$bw) <- "Birth weight, categorized"
Hmisc::label(d$race) <- "Race"
Hmisc::label(d$ht) <- "Arterial hypertension"
Hmisc::label(d$ui) <- "Uterine irritability"
```

## The result {.smaller}

```{r table2, echo = TRUE}
library(compareGroups)
descrTable(smoke ~ bwt + bw + race + ht + ui, data = d,
           show.p.overall = FALSE,
           hide.no = c("no-AHT", "no-UI")) %>%   
  export2md(caption = "Descriptive analysis by smoking status") 

```

## Referencing figures and tables in text

`\ref{tab:chunk_name}` for tables

`\ref{fig:chunk_name}` for figures

\

Examples:  

- If a code chunk name `descriptive_analysis` produces a table, it can be referenced in text as `\ref{tab:descriptive_analysis}`

- If a code chunk name `boxplot_bwt` produces a graphic, it can be referenced in text as `\ref{fig:boxplot_bwt}`

\
Tables and figures will be automatically numbered.

## Inline code

The output of R code can be inserted in a text paragraph with  `` `r knitr::inline_expr("expression")` ``

Example:  

Birth weights ranged from `` `r knitr::inline_expr("min(d$bwt)")` `` to `` `r knitr::inline_expr("max(d$bwt)")` `` grams, with mean (SD) of `` `r knitr::inline_expr("round(mean(d$bwt),1)")` `` (`` `r knitr::inline_expr("round(sd(d$bwt),1)")` ``) grams. 

Becomes:

Birth weights ranged from `r min(d$bwt)` to `r max(d$bwt)` grams, with mean (SD) of `r round(mean(d$bwt),1)` (`r round(sd(d$bwt),1)`) grams. 


## Extracting results from analyses

```{r }
result <- t.test(bwt ~ smoke, data = d)
p <- result$p.value %>% round(3)
mean_ci <- paste0(round(result$estimate[1] - result$estimate[2]),
                  " [95% CI: ",
                  round(result$conf.int)[1],
                  ", ",
                  round(result$conf.int)[2],
                  "]")
```

Birth weights tended to be lower in smoking mothers than in non-smoking mothers (t-test p = `` `r knitr::inline_expr("p")` ``), with a mean difference of `` `r knitr::inline_expr("mean_ci")` `` grams.

Becomes:

Birth weights tended to be lower in smoking mothers than in non-smoking mothers (t-test p = `r p`), with a mean difference of `r mean_ci` grams.


## Exercise

Update [this .Rmd file](https://raw.githubusercontent.com/acobos/CDB_ed2/main/6_3_Rmarkdown_exe_1_solution.Rmd) of the analysis of birth weights:

1. Include a table with descriptive analysis by smoking status.

2. Include references to this table and to the figures in the text refering to them.

3. Use inline code to report:
    + the number of cases analysed (Introduction section).
    + the range, mean and SD of birth weights (Results section).
    
4. Add a paragraph reporting the results of the t-test on the weight of newborns according to smoking status of mothers, using inline code for the $p$ value and the mean difference and its 95% CI.

5. At each step use the `knit` button to verify.