---
title: "Course index"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)

```

\


```{r}
library(dplyr)
library(kableExtra)

options(knitr.kable.NA = '')

rio::import("teaching_plan.xlsx", sheet = "index") %>% 
  filter(!is.na(session)) %>% 
#  filter(topic_num == 2)  %>% 
  mutate(slides = paste0("[", 
                         slides_title, 
                         "](",
                         paste0(slides_filename, ifelse(topic_num == 1, ".R", ".html")), ")"
         ),
         exe_solution = paste0("[R script](", slides_filename, "_exe_solution.R)"),
         exe_solution = ifelse(exes == "no", "", exe_solution)
         ) -> d

# pull(d, topic) %>% unique() 


topic_tbl <- function(topic_name){
  
  d %>% 
  filter(topic == topic_name) %>% 
  select(Materials = slides, `Exercise solution` = exe_solution) %>% 
  knitr::kable() %>% 
  kable_styling(full_width = FALSE, position = "left", 
                bootstrap_options = "condensed") %>% 
    column_spec(1, width = "30em") %>% 
    column_spec(2, width = "10em")
}

```


## Software

1. Install R from [CRAN](https://cran.r-project.org/) (choose appropriate version).

2. Install [RStudio](https://www.rstudio.com/products/rstudio/download/), free Desktop version.

3. Open RStudio.

\


```{r}
downloadthis::download_file(
      path = "data.zip",
      button_label = "Download example datasets",
      button_type = "primary",
      has_icon = TRUE,
      icon = "fa fa-save",
      self_contained = FALSE)
```

\

## R language basics
`r topic_tbl("R language basics")`

\

## Working with dataframes
`r topic_tbl("Working with dataframes")`

\

## Graphics
`r topic_tbl("Graphics")`

\

## Comparing groups
`r topic_tbl("Comparing groups")`

\

## Predictions
`r topic_tbl("Predictions")`

\

## Miscellanea
`r topic_tbl("Miscellanea")`

