---
title: "Course index"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)

```

\


```{r}
library(dplyr)
library(kableExtra)

options(knitr.kable.NA = '')

rio::import("teaching_plan.xlsx", sheet = "index") %>% 
  filter(!is.na(session)) %>% 
#  filter(topic_num == 2)  %>% 
  mutate(slides = paste0("[", 
                         slides_title, 
                         "](",
                         paste0(slides_filename, ifelse(topic_num == 1, ".R", ".html")), ")"
         ),
         exe_solution = paste0("[R script](", slides_filename, "_exe_solution.R)"),
         exe_solution = ifelse(exes == "no", "", exe_solution)
         ) -> d

# pull(d, topic) %>% unique() 

d$exe_solution <- ifelse(d$slides_title == "Rmarkdown",
                         "[report.Rmd](https://raw.githubusercontent.com/acobos/CDB_ed2/main/6_3_Rmarkdown_exe_1_solution.Rmd),  [slides.Rmd](https://raw.githubusercontent.com/acobos/CDB_ed2/main/6_3_Rmarkdown_exe_2_solution.Rmd)",
                         d$exe_solution)

d$exe_solution <- ifelse(d$slides_title == "More Rmarkdown",
                         "[report.Rmd](https://raw.githubusercontent.com/acobos/CDB_ed2/main/6_4_more_Rmarkdown_exe_solution.Rmd)",
                         d$exe_solution)


topic_tbl <- function(topic_name){
  
  d %>% 
  filter(topic == topic_name) %>% 
  select(Materials = slides, `Exercise solution` = exe_solution) %>% 
  knitr::kable() %>% 
  kable_styling(full_width = FALSE, position = "left", 
                bootstrap_options = "condensed") %>% 
    column_spec(1, width = "30em") %>% 
    column_spec(2, width = "20em")
}

```


## Software

1. Install R from [CRAN](https://cran.r-project.org/) (choose appropriate version).

2. Install [RStudio](https://www.rstudio.com/products/rstudio/download/), free Desktop version.

3. Open RStudio.

\


```{r}
downloadthis::download_file(
      path = "data.zip",
      button_label = "Download example datasets",
      button_type = "primary",
      has_icon = TRUE,
      icon = "fa fa-save",
      self_contained = FALSE)
```

\

## R language basics
`r topic_tbl("R language basics")`

\

## Working with dataframes
`r topic_tbl("Working with dataframes")`

- [Cheat sheet on Data Transformation with dplyr](https://nyu-cdsc.github.io/learningr/assets/data-transformation.pdf)

- [RStudio cheat sheets collection](https://www.rstudio.com/resources/cheatsheets/)



\

## Graphics
`r topic_tbl("Graphics")`

- [The R graph gallery](https://r-graph-gallery.com/)

- No need to code with [ggplotgui](https://shiny.gmw.rug.nl/ggplotgui/), but you can get the R code!  

\

## Comparing groups
`r topic_tbl("Comparing groups")`

- [A Student’s guide to R](https://cran.r-project.org/doc/contrib/Horton+Pruim+Kaplan_MOSAIC-StudentGuide.pdf): free book on introductory and intermediate statistics with R.

- Do not use histograms to assess normality: [find out why](https://statisticsbyjim.com/basics/assessing-normality-histograms-probability-plots/).

- For an intuitive introduction to the bootstrap, see [this](https://statisticsbyjim.com/hypothesis-testing/bootstrapping/).

\

## Predictions
`r topic_tbl("Predictions")`

- [Survival analysis/plots cheat sheet](https://rpkgs.datanovia.com/survminer/survminer_cheatsheet.pdf)

- [Regression modeling strategies](https://www.amazon.com/Regression-Modeling-Strategies-Applications-Statistics/dp/0387952322) is a _very_ good book on modeling (with R).

\

## Miscellanea
`r topic_tbl("Miscellanea")`

- [Packages for ROC curves](https://rviews.rstudio.com/2019/03/01/some-r-packages-for-roc-curves/)

- [Rmarkdown cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)